ppseq: An R Package for Sequential Predictive Probability Monitoring

### 2022-09-28 inst/ppseq-package-paper/zabor-hobbs-kane-1-review-response.Rmd

---
title: "ppseq: An R Package for Sequential Predictive Probability Monitoring"
subtitle: "Review 1 Responses"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

> **Reviewer 1**

> Overview:
> The authors present an R package to design early phase clinical trials with binary outcomes leveraging a Bayesian framework. I find the package potentially applicable to many settings. However, the paper includes some statements that are not sufficiently clear, or debatable.  Some details should be clarified. I also think that a running toy-example that can be implemented without the use of a server would be a good tutorial to familiarize with the package before large scale implementation. Some comments to follow:

Thank you for this thorough review. We agree that there were parts of the paper that were unclear or lacked sufficient detail, and have strived to address all of your comments in an effort to improve the clarify and accuracy throughout.

> Major comments:

> - The package and paper only consider binary outcomes. This should be clearly stated in the abstract/introduction.

Thank you for pointing this out. We agree, and have added this detail to both the abstract and the last paragraph of the introduction.

> - I am not sure what is the point of second paragraph of the introduction. In particular, the authors claim that on-the fly decisions can lead to “large sample size” and “poor statistical properties”. Why? I don’t think that this is necessarily the case. I suggest improving the clarity of this paragraph.

The goal of this paragraph was to introduce readers to the often large sample sizes that have been used in recent expansion cohorts, which is part of the problem addressed by this package. We see how the wording did not emphasize this point properly, though, and have reworded in response to this comment. Thank you.

> - In the third paragraph of the introduction the author claims that Bayesian models are flexible “while also maintaining the control of type-I error”. I disagree with this statement. Bayesian models do not naturally account for type-I error and power (frequentist power?) constraints , which is the whole point of the calibration exercise that motivate this package.

We agree that the way this sentence was worded was not accurate. We have reworded this section to improve clarity and accuracy.

> - In the introduction and later in the paper, the authors introduce “one-sample” and “two-sample” phase-I design. The meaning should be explained somewhere.

We removed this language from the introduction, and added a definition of one-sample and two-sample to the beginning of the "Predictive probability monitoring" section. In this context, one-sample refers to a trial in which all patients are enrolled onto a single arm and given the experimental treatment and two-sample refers to a trial in which patients are enrolled onto two different treatment arms, or a treatment and control arm, for comparative purposes.

> - Page 3: “The proportion of positive trials under the null represents the type I error and the proportion of positive trials under the alternative represents the power.” I don’t think this is correct. Specifically, I am not sure what “represent” means in this context.  The proportion of “positive trial” (maybe better defined as the proportion of simulated trials for which we reject the null hypothesis of inferiority, or something along this line) is a Monte Carlo estimate of the type-I error. Correct? Same goes for power. The number of trials replications should also be discussed somewhere in the paper. I think this is very important aspect   when selecting an optimal strategy, i.e., we don’t want the monte Carlo error to dominate the difference between alternative strategies. This aspect should also be discussed (see also following point).

Thank you for pointing out the lack of specificity in our wording. We meant to convey that the proportion of trials simulated under the null hypothesis for which we reject the null hypothesis is an estimate of the type I error. And similarly the proportion of trials simulated under the alternative hypothesis for which we reject the null hypothesis is an estimate of the power. We have revised the sentence accordingly.

And you are correct that the number of trial replications is an important aspect of this. Our default arguments are 1000 datasets with 5000 posterior draws in each, which we believe are reasonable values. We have added a note of caution related to these values in the Package Overview section.

> - I find the section “Optimization” that describe the function “optimal_design()” to be misleading. First, I think that the authors should explicitly state that this function does not perform an optimization, but rather select a trial among the ones that have been simulated. There is no guarantee that the selected design is in fact optimal with respect to any criteria (e.g., if we select a coarse grid of thresholds). Second, I think that with a minimal effort this function can be generalized to take as input any user specified utility that take as impute type-I error and power, and/or average sample sizes. I think this could be helpful for a broader use of the package.

This is a very good point, and we apologize for the poor word choice here. We have renamed this section "Design selection" and have removed the word optimization to avoid any confusion. Since the paper was originally drafted, we have added the option of weighting the various criteria, to allow more user-specified flexibility in terms of the value of type I error versus power, and average sample size under the null versus average sample size under the alternative. We have added a description of this new functionality to the Design selection section.

> - In the case study: what is the rational to select the considered posterior thresholds and not, say, an equi-spaced grid. I think this is an important point to discuss, and can have a large impact on the result.

Thank you, we have written the case study a bit to be a two-sample example in order to differentiate from the added basic example, and have also added some information justifying the selected thresholds.

> - In the code snipped at page 4 the parameters `S` and `nsim` of the `calibrate_thresholds()` function have not been discussed in the text. I think this are important parameters.

We have added more discussion of the importance of selecting sufficiently large values of `S` and `nsim` throughout the paper, thank you.

> - Figure 1: I think the x-axis should be rescaled in a (say 0-0.25 for the left panel and 35-55 for the right panel). The plot would be easier to read.

Thank you for this suggestion, we have rescaled the axes.

> - Page 7: In the first line of the tibble `one_sample_decision_tbl` there is NA,NA. If I am not mistaken this is interpreted as “never stop”. I will explicitly indicate this.
> - There are some typos in the bibliography. For example, “r” in place of “R”, “phase i” in place of “phase I”. Please double check.
> - I think that the paper would benefit of an additional toy example that can be run on a personal computer rather than on a server. This would allow to familiarize with the functions before running large scale simulations.

We agree, and have added a toy example that can easily be run interactively prior to presenting the more realistic case study.

> Minor comments:

> - The third sentence in the abstract “Recent successes […]” is very long  and convoluted. I suggest rephrasing.

We have taken this suggestion and revised the abstract to improve clarity.

> - Page 2: Define the acronym RECIST.

In parentheses we have added the definition of the acronym RECIST: Response Evaluation Criteria in Solid Tumors.

> - Page 2: Bayes rule -> Bayes’s theorem.

Thank you, we have made this correction.

> - Page2: when naming distribution (e.g., Beta-Binomial) I suggest to use $\mbox{Beta-Binomial}()$ or similar, to avoid wrong spacing/font.

Thank you for this suggestion, we have used it for all distributions to achieve the proper font/spacing.

> - In the section decision rules, what is a decision point “r”? Please define this.

We have reworded this for clarity, thank you.

> -Figure 1: in the legend (#fig:unnamed-chunk-7). I suppose this is a compilation error.

This is indeed a compilation error. I am not sure why this is being generated or how to suppress it, so hopefully it can be fixed when the article is finalized for publication,

> **Reviewer 5**

> The article describes the ppseq package for R, which provides several functions to facilitate clinical trial design using sequential predictive probability monitoring. A user looking to design an early phase trial can use Bayesian methods to ask what the probability of early stopping is and what expected sample sizes would be under different assumptions about outcomes in treatment and control groups and with different stopping rules, while maintaining desired type I error rates. The package is small, providing only three user-facing functions reported in the manuscript, along with plot and print methods. However, the functions are user friendly with obvious options, and the article provides a clear use case and justification for the methods. However, the article itself is quite difficult to follow, which is not helped by the example not being feasible to run for a user with a standard desktop or laptop computer (the example is reportedly run here on a 192 core computing cluster!). I have provided some queries, recommendations below.

We are sorry that the article in it's original form was difficult to follow. We hope the changes have made it more understandable, and that all of the needed detail is now included. Thank you for taking the time to conduct such a thorough review of our paper, we do believe that it has been improved as a result.

> 1. I found the Package Overview section difficult to follow, since it provides only a narrative description of the inputs of the functions and options. It would be useful to link the description to the statistical discussion in the previous section, explaining which option relates to which parameter, and what each output reports in terms of the statistical notation. It may help to integrate the case study into the package description so that the reader can see what the text means in terms of the example. However, I would also strongly recommend providing an example(s) that a user can implement themselves in a reasonable amount of time, as this often helps when using a package for the first time. Indeed, is the package useful if it requires such significant computational resources (see also point 3)?

Thank you for this comment, we agree that much detail was overlooked in the Package Overview section, and we have expanded this section to include much more detail of the underlying algorithm, and to link the notation to the preceding section. We believe this is much easier to follow now. We have also added a section for a toy example that can easily be run interactively prior to presenting the more detailed case study.

We do think this package is useful despite the potentially long computational time. This is meant to be used at the design stage of a clinical trial, when there is ample time to allow for computation. Once the design has been selected and decided, no additional computations are required throughout the conduct of the trial, as the decision table resulting from `calc_decision_rules()` is available for decisions at each interim look, making trial conduct fast at seamless at each interim analysis and at the end of the trial.

> 2. I was a little confused by the notation, some more detail would make it clearer I believe. Is the setting a randomized trial with treatment and control groups, or a dose expansion cohort? We are told that $p$ is the probability of response – is this in the dose expansion cohort? And we have $p_0$ and $p_1$ as the probabilities of response in the control and treatment groups, respectively. But are these random variables we intend to estimate, or are they pre-defined thresholds of interest (e.g. “high” and “low” efficacy)?

Thank you for pointing out the lack of clarity here. We have revised the wording to clarify that $p$ represents the true response probability, $p_0$ is a fixed threshold of unacceptable response rate, $p_1$ is a fixed threshold of acceptable response rate. We have also clarified that our example focuses on the setting of a single arm dose-expansion cohort.

> 3. It was not clear to me what each function was doing computationally. In particular, the method implemented by the `calibrate_thresholds()` function was not explicated. I wanted to understand why the function took so long to run, and why it might be drawing posterior samples. My assumption was that the algorithm was something roughly like the following. For nsim iterations:
i. Simulate a realization of $p^r$ from the prior $\Beta(a_0, b_0)$
ii. Simulate a realization of $x^r | n, p^r ~ \Binomial(n, p^r)$
iii. Calculate the probability of success in that trial $$PP^r=\Pr(p^r>p_0|x^r, N, n)=\int_{p_0}^1 f_p(q|x^r, N, n)\,dq=\sum_{x^*}^{N-n}\int_{p_0}^1\Pr(X^*=x^*|q,N-n)f_p(q|x^r,N,X^*=x^*)\,dq

> Then the PPP is the the proportion of $PP^r>\theta^*$. If this is the case, then the integral term has a closed form and posterior simulations do not seem necessary, in which case the function should run much quicker than it currently does. Running time does seem like a key issue for the functions in this package. I may of course be completely wrong about what the function is doing, but it would help to have some clarification so we know the function is doing what the user expects.

Thank you for drawing our attention to this oversight. We have added much more detail about the underlying computational algorithm for the `calibrate_thresholds()` function in the Package Overview section, and hope that this clarifies the functionality.

>4. The section “Optimisation” mentions minimizing Euclidean distance on different plots, however it isn’t clear what this means without notation or until the example is provided.

Thank you for pointing out the lack of clarity in this section. We have added much more detail about the computations being done, with notation, to this section, and hope it is now more clear. Additionally, in response to another reviewer's comment we have renamed this section "Design selection" and removed reference to optimization, since no true mathematical optimization was being conducted.

### 2022-09-28 inst/ppseq-package-paper/zabor-hobbs-kane.Rmd
drafting revision 1 response

---
title: "ppseq: An R Package for Sequential Predictive Probability Monitoring"
abstract: >
  Advances in drug discovery have produced numerous biomarker-guided therapeutic strategies for treating cancer. Yet the promise of precision medicine comes with the cost of increased complexity. Recent successes for non-cytotoxic treatments have relied on expansive early phase clinical trials that far exceeded the size of trials typically implemented for chemotherapies, with the addition of baskets for different disease sub-types or multiple dosing groups, expansion cohorts to further study safety and obtain preliminary efficacy information, or randomization to further define efficacy or compare across doses, among other design elements. With the enlarged sample sizes come increasing ethical concerns regarding patients who enroll in clinical trials, and the need for rigorous statistical designs to ensure that trials can stop early for futility while maintaining traditional control of type I error and power. The R package \pkg{ppseq} provides a framework for designing early phase clinical trials using sequential futility monitoring based on the Bayesian predictive probability. Trial designs can be compared using interactive plots and optimized based on efficiency or accuracy.
draft: true
author:  
  # see ?rjournal_article for more information
  - name: Emily C. Zabor
    affiliation: Department of Quantitative Health Sciences & Taussig Cancer Institute, Cleveland Clinic
    address:
    - 9500 Euclid Ave. CA-60
    - Cleveland, OH 44195 USA
    url: http://www.emilyzabor.com/
    orcid: 0000-0002-1402-4498
    email:  zabore2@ccf.org
  - name: Brian P. Hobbs
    email:  brian.hobbs@austin.utexas.edu
    orcid: 0000-0003-2189-5846
    affiliation: Dell Medical School, The University of Texas at Austin
    address:
    - 1501 Red River Street MC: Z0100
    - Austin, TX 78712
  - name: Michael J. Kane
    email: michael.kane@yale.edu
    orcid: 0000-0003-1899-6662
    affiliation: Department of Biostatistics, Yale University
    address:
    - 60 College Street
    - New Haven, CT 06511
date: "August 20, 2021"
type: package
creative_commons: CC BY
output: 
  rjtools::rjournal_web_article
bibliography: zabor-hobbs-kane.bib
---

@@

In the context of cytotoxic treatments, phase 1 trials in oncology traditionally have the primary aim of identifying the maximum tolerated dose (MTD), defined as the highest dose that still maintains a certain pre-specified rate of toxicity, most commonly 30\%, in a dose-escalation phase. Designs for dose-escalation trials include the rule-based 3+3 design and the model-based continual reassessment method, among others. But with increasing study focused on non-cytotoxic treatments, such as immunotherapies, the MTD either may not exist or may not be relevant, and toxicities may either develop much later or even be chronic, making these treatments difficult to study with traditional dose-escalation designs [@Pestana2020]. As a result, it is becoming increasingly common to include dose-expansion cohorts, in which additional patients are enrolled in phase 1 after the dose-escalation phase is complete. In this setup, the dose-escalation phase is considered phase 1a and used to assess the initial safety of multiple doses, then the dose-expansion phase is considered phase 1b and can have a variety of aims including to further refine the safety of one or more doses, to assess preliminary efficacy, to explore the treatment in various disease-specific subtypes that all share a common biomarker that the treatment is targeting, or to further characterize the pharmacokinetics and/or pharmacodynamics. The use of dose-expansion cohorts increased from 12\% in 2006 to 38\% in 2011 [@Manji2013] and trials with dose-expansion cohorts led to higher response rates and more frequent success in phase 2 trials [@Bugano2017].

But dose-expansion cohorts are not always planned in advance, and therefore may be subject to on-the-fly decision making that can lead to large sample sizes and poor statistical properties. For example, the KEYNOTE-001 trial of pembrolizumab, initially designed as a 3+3 dose-escalation trial, included multiple protocol amendments and ultimately enrolled a total of 655 patients across five melanoma expansion cohorts and 550 patients across four non-small-cell lung cancer expansion cohorts [@Khoja2015]. In a basket trial of atezolizumab, an anti-PD-L1 treatment in patients with a variety of cancers and both with and without PD-L1 expression, an expansion cohort in metastatic urothelial bladder cancer ultimately enrolled 97 patients and evaluated 95, despite the fact that no expansion cohort in this disease subtype was originally planned in the trial protocol. The expansion cohort in metastatic urothelial carcinoma was rather added later in a protocol amendment in which the sample size was increased from what was initially planned [@Petrylak2018; @Powles2014].

Bayesian sequential predictive probability monitoring provides a natural framework for early phase oncology trial design, allowing flexibility to stop early for futility or safety concerns while also maintaining traditional levels of power and control of type I error. As a result, predictive probability has been proposed as an approach to interim monitoring in clinical trials previously [@Dmitrienko2006; @Lee2008; @Saville2014; @Hobbs2018]. However, in order to be useful to investigators designing such trials, software must be made available to select an optimal design. This paper introduces the \CRANpkg{ppseq} package for the R software language [@RCT2020], which provides functions to design one-sample and two-sample early phase clinical trials using sequential predictive probability monitoring for futility. Interactive plots produced using the \CRANpkg{ggplot2} package [@Wickham2016] and the \CRANpkg{plotly} package [@Sievert2020] compare designs based on different thresholds for decision making. Moreover, we demonstrate optimization criteria for selecting an ideal predictive probability monitoring design using the \CRANpkg{ppseq} package. While the \CRANpkg{ppseq} package was developed with early phase oncology clinical trials in mind, the methodology is general and can be applied to any application of sequential futility monitoring in clinical trial design.

@@

Consider the setting of a binary outcome, such as tumor response as measured by the RECIST criteria in the setting of a study in patients with solid tumors, where each patient, denoted by $i$, enrolled in the trial either has a response such that $x_i = 1$ or does not have a response such that $x_i = 0$. Then $X = \sum_{i=1}^n x_i$ represents the number of responses out of $n$ currently observed patients up to a maximum of $N$ total patients. Let $p$ represent the probability of response, where $p_0$ represents the null response rate under no treatment or the standard of care treatment and $p_1$ represents the alternative response rate under the experimental treatment. Most dose-expansion studies with an efficacy aim will wish to test the null hypothesis $H_0: p \leq p_0$ versus the alternative hypothesis $H_1: p \geq p_1$.

The Bayesian paradigm of statistics is founded on Bayes rule, which is a mathematical theory that specifies how to combine the prior distributions that define prior beliefs about parameters, such as the response rate $p$, with the observed data, such as the total number of responses $X$, yielding a posterior distribution. Here the prior distribution of the response rate $\pi(p)$ has a beta distribution $Beta(a_0, b_0)$ and our data $X$ have a binomial distribution $bin(n, p)$. Combining the likelihood function for the observed data $L_x(p) \propto p^x (1-p)^{n-x}$ with the prior, we obtain the posterior distribution of the response rate, which follows the beta distribution $p|x \sim Beta(a_0 + x, b_0 + n - x)$. A posterior probability threshold $\theta$ would be pre-specified during the trial design stage. At the end of the trial if the posterior probability exceeded the pre-specified threshold, i.e. if $\Pr(p>p_0 | X) > \theta$, the trial would be declared a success.

The posterior predictive distribution of the number of future responses $X^*$ in the remaining $n^*=N-n$ future patients follows a beta-binomial distribution $Beta-binomial(n^*, a_0 + x, b_0 + n - x)$. Then the posterior predictive probability (PPP), is calculated as $PPP = \sum_{{x^*}=0}^{n^*} \Pr(X^*=x^*|x) \times I(\Pr(p>p_0 | X, X^*=x^*) > \theta)$. The posterior predictive probability represents the probability that, at any given interim monitoring point, the treatment will be declared efficacious at the end of the trial when full enrollment is reached, conditional on the currently observed data and the specified priors. We would stop the trial early for futility if the posterior predictive probability dropped below a pre-specified threshold $\theta^*$, i.e. $PPP<\theta^*$. Predictive probability thresholds closer to 0 lead to less frequent stopping for futility whereas thresholds near 1 lead to frequent stopping unless there is almost certain probability of success. Predictive probability provides an intuitive interim monitoring strategy for clinical trials that tells the investigator what the chances are of declaring the treatment efficacious at the end of the trial if we were to continue enrolling to the maximum planned sample size, based on the data observed in the trial to date.

@@

The \CRANpkg{ppseq} package facilitates the design of clinical trials utilizing sequential predictive probability monitoring for futility. The goal is to establish a set of decision rules at the trial planning phase that would be used for interim monitoring during the course of the trial. The main challenge in designing such a trial is joint calibration of the posterior probability and posterior predictive probability thresholds to be used in the trial in order to maintain the desired level of power and type I error. The `calibrate_thresholds()` function will evaluate a grid of posterior and predictive thresholds provided by the user as vector inputs through the argument `pp_threshold` for posterior thresholds and the argument `ppp_threshold` for predictive thresholds. Other required arguments include the null response rate `p_null`, the alternative response rate `p_alt`, a vector of sample sizes at which interim analyses are to be performed `n` as well as the maximum total sample size `N`. Default arguments are available for the direction of the alternative hypothesis `direction`, a vector of the two hyperparameters of the prior beta distribution `prior`, and the number of posterior samples `S` and the number of simulated trial datasets `nsim`, but customization is also possible. The additional argument `delta` can specify the clinically meaningful difference between groups in the case of a two-sample trial design. The function returns a list, the first element of which is a `tibble` containing the posterior threshold, predictive threshold, the mean sample size under the null and the alternative, the proportion of positive trials under the null and alternative, and the proportion of trials stopped early under the null and alternative. The proportion of positive trials under the null represents the type I error and the proportion of positive trials under the alternative represents the power. The `print()` option will print the results summary for each combination of thresholds, filtered by an acceptable range of type I error and minimum power, if desired.

## Optimization

After obtaining results for all combinations of evaluated posterior and predictive thresholds, the next step is to select the ideal design from among the various options. The \CRANpkg{ppseq} package introduces two optimization criteria to assist users in making a selection. The first, called the "optimal accuracy" design, identifies the design that minimizes the Euclidean distance to the top left point on a plot of the type I error by the power. The second, called the "optimal efficiency" design, identifies the design that minimizes the Euclidean distance to the top left point on a plot of the average sample size under the null by the average sample size under the alternative, subject to constraints on the type I error and power. The `optimize_design()` function will return a list that contains the details of each of the two optimal designs.

@@

plan(multicore(workers = 76))

>>>>

---
title: "ppseq: An R Package for Sequential Predictive Probability Monitoring"
abstract: >
  Advances in drug discovery have produced numerous biomarker-guided therapeutic strategies for treating cancer. Yet the promise of precision medicine comes with the cost of increased complexity. Recent trials of targeted treatments have included expansion cohorts with sample sizes far exceeding those in traditional early phase trials of chemotherapeutic agents. The enlarged sample sizes raise ethical concerns for patients who enroll in clinical trials, and emphasize the need for rigorous statistical designs to ensure that trials can stop early for futility while maintaining traditional control of type I error and power. The R package \pkg{ppseq} provides a framework for designing early phase clinical trials of binary endpoints using sequential futility monitoring based on Bayesian predictive probability. Trial designs can be compared using interactive plots and selected based on measures of efficiency or accuracy.
draft: true
author:  
  # see ?rjournal_article for more information
  - name: Emily C. Zabor
    affiliation: Department of Quantitative Health Sciences & Taussig Cancer Institute, Cleveland Clinic
    address:
    - 9500 Euclid Ave. CA-60
    - Cleveland, OH 44195 USA
    url: http://www.emilyzabor.com/
    orcid: 0000-0002-1402-4498
    email:  zabore2@ccf.org
  - name: Brian P. Hobbs
    email:  brian.hobbs@austin.utexas.edu
    orcid: 0000-0003-2189-5846
    affiliation: Dell Medical School, The University of Texas at Austin
    address:
    - 1501 Red River Street MC: Z0100
    - Austin, TX 78712
  - name: Michael J. Kane
    email: michael.kane@yale.edu
    orcid: 0000-0003-1899-6662
    affiliation: Department of Biostatistics, Yale University
    address:
    - 60 College Street
    - New Haven, CT 06511
date: "August 20, 2021"
type: package
creative_commons: CC BY
output: 
  rjtools::rjournal_web_article
bibliography: zabor-hobbs-kane.bib
---

@@

Statistical methods for early phase oncology trials were developed in the context of cytotoxic treatments. Most cytotoxic treatments exhibit increases in both efficacy and toxicity with increasing dose. As a result, phase I trials were centered on identifying the maximum tolerated dose (MTD), defined as the highest dose that did not exceed a pre-specified toxicity threshold. Phase I dose-escalation trials were designed using the rule-based 3+3 method or the model-based continual reassessment method, among others. But advances in drug discovery have produced numerous biomarker-guided non-cytotoxic therapies such as small molecule inhibitors, antibody drug conjugates, immune checkpoint inhibitors, and monoclonal antibodies. These therapies typically do not exhibit the same pattern of increases in both efficacy and toxicity with increasing dose, so the MTD as traditionally defined may not exist. Instead, lower doses may have equivalent efficacy but lower toxicity as compared to higher doses. As a result, these therapies can be difficult to study with traditional dose-escalation designs [@Pestana2020]. To address this issue, recent phase I trials have included large dose-expansion cohorts, in which additional patients are enrolled in phase 1 after the dose-escalation phase is complete. In this setup, the dose-escalation phase is considered phase 1a and used to assess the initial safety of multiple doses, then the dose-expansion phase is considered phase 1b and can have a variety of aims including to further refine the safety of one or more doses, to assess preliminary efficacy, to explore the treatment in various disease-specific subtypes, or to further characterize the pharmacokinetics and/or pharmacodynamics. The use of dose-expansion cohorts increased from 12\% in 2006 to 38\% in 2011 [@Manji2013] and trials with dose-expansion cohorts led to higher response rates and more frequent success in phase 2 trials [@Bugano2017].

But despite these successes, recent dose-expansion cohorts have not always been planned in advance, leading to uncertain statistical properties, and have at times included samples sizes that far exceed those typically seen in early phase trials of cytotoxic treatments. For example, the KEYNOTE-001 trial of pembrolizumab, initially designed as a 3+3 dose-escalation trial, included multiple protocol amendments and ultimately enrolled a total of 655 patients across five melanoma expansion cohorts and 550 patients across four non-small-cell lung cancer expansion cohorts [@Khoja2015]. In a basket trial of atezolizumab, an anti-PD-L1 treatment in patients with a variety of cancers and both with and without PD-L1 expression, an expansion cohort in metastatic urothelial bladder cancer ultimately enrolled 97 patients and evaluated 95, despite the fact that no expansion cohort in this disease subtype was originally planned in the trial protocol. The expansion cohort in metastatic urothelial carcinoma was rather added later in a protocol amendment in which the sample size was increased from what was initially planned [@Petrylak2018; @Powles2014]. These enlarged sample sizes raise ethical concerns for patients who enroll in clinical trials, and emphasize the need for rigorous statistical designs to ensure that trials can stop early for futility while maintaining traditional control of type I error and power.

Bayesian predictive probability has been proposed as an approach for sequential monitoring in early phase oncology trials [@Dmitrienko2006; @Lee2008; @Hobbs2018; @Saville2014]. However, in order to be useful to investigators designing such trials, software must be made available to calibrate the design for the desired statistical properties. To our knowledge no such software currently exists in the R programming language. This paper introduces the \CRANpkg{ppseq} package for the R software language [@RCT2020], which provides functions to design early phase clinical trials of binary endpoints using sequential predictive probability monitoring for futility. Interactive plots produced using the \CRANpkg{ggplot2} package [@Wickham2016] and the \CRANpkg{plotly} package [@Sievert2020] compare designs based on different thresholds for decision making. Moreover, we demonstrate criteria for selecting an ideal predictive probability monitoring design using the \CRANpkg{ppseq} package. While the \CRANpkg{ppseq} package was developed with early phase oncology clinical trials in mind, the methodology is general and can be applied to any application of sequential futility monitoring in clinical trial design.

@@

Consider the setting of an expansion cohort with a binary outcome, such as tumor response as measured by the RECIST (Response Evaluation Criteria in Solid Tumors) criteria. Here we will focus on the one-sample setting, in which all patients in the trial are enrolled onto a single arm and given the experimental treatment of interest. Functionality is also available for the two-sample setting, in which patients are enrolled onto two different treatment arms, or a treatment and a control arm, for comparative purposes. Each patient, denoted by $i$, enrolled in the trial either has a response such that $x_i = 1$ or does not have a response such that $x_i = 0$. Then $X = \sum_{i=1}^n x_i$ represents the number of responses out of $n$ currently observed patients up to a maximum of $N$ total patients. Let $p$ represent the true probability of response. Fix $p_0$ as the threshold for unacceptable response rate and $p_1$ as the threshold for acceptable response rate. Most dose-expansion studies with an efficacy aim will wish to test the null hypothesis $H_0: p \leq p_0$ versus the alternative hypothesis $H_1: p \geq p_1$.

The Bayesian paradigm of statistics is founded on Bayes' theorem, which is a mathematical theory that specifies how to combine the prior distributions that define prior beliefs about parameters, such as the true response rate $p$, with the observed data, such as the total number of responses $X$, yielding a posterior distribution. Here the prior distribution of the response rate $\pi(p)$ has a beta distribution $\mbox{Beta}(a_0, b_0)$ and our data $X$ have a binomial distribution $\mbox{Bin}(n, p)$. Combining the likelihood function for the observed data $L_x(p) \propto p^x (1-p)^{n-x}$ with the prior, we obtain the posterior distribution of the response rate, which follows the beta distribution $p|x \sim \mbox{Beta}(a_0 + x, b_0 + n - x)$. A posterior probability threshold $\theta$ would be pre-specified during the trial design stage. At the end of the trial if the posterior probability exceeded the pre-specified threshold, i.e. if $\Pr(p>p_0 | X) > \theta$, the trial would be declared a success.

The posterior predictive distribution of the number of future responses $X^*$ in the remaining $n^*=N-n$ future patients follows a beta-binomial distribution $\mbox{Beta-binomial}(n^*, a_0 + x, b_0 + n - x)$. Then the posterior predictive probability (PPP), is calculated as $PPP = \sum_{{x^*}=0}^{n^*} \Pr(X^*=x^*|x) \times I(\Pr(p>p_0 | X, X^*=x^*) > \theta)$. The posterior predictive probability represents the probability that, at any given interim monitoring point, the treatment will be declared efficacious at the end of the trial when full enrollment is reached, conditional on the currently observed data and the specified priors. We would stop the trial early for futility if the posterior predictive probability dropped below a pre-specified threshold $\theta^*$, i.e. $PPP<\theta^*$. Predictive probability thresholds closer to 0 lead to less frequent stopping for futility whereas thresholds near 1 lead to frequent stopping unless there is almost certain probability of success. Predictive probability provides an intuitive interim monitoring strategy for clinical trials that tells the investigator what the chances are of declaring the treatment efficacious at the end of the trial if we were to continue enrolling to the maximum planned sample size, based on the data observed in the trial to date.

@@

The \CRANpkg{ppseq} package facilitates the design of clinical trials utilizing sequential predictive probability monitoring for futility. The goal is to establish a set of decision rules at the trial planning phase that would be used for interim monitoring during the course of the trial. The main computational challenge in designing such a trial is joint calibration of the posterior probability and posterior predictive probability thresholds to be used in the trial in order to achieve the desired levels of frequentist type I error and power. The main function for achieving this aim is the `calibrate_thresholds()` function, which will evaluate a grid of posterior thresholds $\theta$ and predictive thresholds $\theta^*$ provided by the user as vector inputs specified with the arguments `pp_threshold` and `ppp_threshold`, respectively. Other required arguments include the unacceptable response rate $p_0$ specified by `p_null`, the acceptable response rate $p_1$ specified by `p_alt`, a vector of sample sizes at which interim analyses are to be performed `n`, and the maximum total sample size `N`. The direction of the alternative hypothesis is specified with the argument `direction` and defaults to `"greater"`, which corresponds to the alternative hypothesis $H_1: p \geq p_1$. The hyperparameters of the prior beta distribution are specified with the argument `prior` and default to `c(0.5, 0.5)`, which denotes a $\mbox{Beta}(0.5, 0.5)$ distribution. The number of posterior samples are specified with the argument `S`, which defaults to `5000` samples, and the number of simulated trial datasets is specified with the argument `nsim`, which defaults to `1000`. The additional argument `delta`, which defaults to `NULL` for the one-sample setting, can specify a clinically meaningful difference between groups $\delta = p_1 - p_0$ in the case of a two-sample trial design.

The `calibrate_thresholds()` function conducts the following algorithm, given the default arguments:

1. Generate `nsim` datasets, denoted by $j$, containing the cumulative number of responses $x$ at each interim sample size $n$ from a binomial distribution under the unacceptable (i.e. null) response rate $p_0$, specified by `p_null`, and under the acceptable (i.e. alternative) response rate $p_1$, specified by `p_alt`, for each interim look, denoted by $l$. 
2. For dataset $j$ and posterior threshold $\theta_k$, draw `S` samples, denoted by $s$, from the posterior distribution $p|x_{ls} \sim \mbox{Beta}(a_0 + x_l, b_0 + n_l - x_l)$ where $x_l$ is the number of responses at interim look $l$ and $n_l$ is the number of enrolled patients at interim look $l$. Use each $p|x_{ls}$ as the response probability in a binomial distribution to generate the number of future responses $X^*_{ls}$ in the remaining $n^*_l=N-n_l$ future patients at interim look $l$.

    a. Then for each $X^*_{ls}$, generate `S` posterior probabilities, denoted by $s'$, at the end of the trial: $PP^*_{lss'} \sim \mbox{Beta}(a_0 + X^*_{ls} + x_l, b_0 + n^*_l - (X^*_{ls} + x_l))$ and calculate $PP^*_{ls} = \Pr(p>p_0 | X^*_{ls}) = \frac{\sum_1^{S'} PP^*_{lss'} > p_0}{S'}$.
    b. Estimate the predictive probability at posterior threshold $k$ as $PPP_{lk} = \frac{\sum_1^S PP^*_{ls}>\theta_k}{S}$.
    c. Stop the trial for dataset $j$ at interim look $l$ and predictive threshold $m$ if $PPP_{lk} < \theta^*_m$. Otherwise continue enrolling.
3. Repeat (2) over all combinations of datasets $j$, posterior thresholds $k$, and predictive thresholds $m$.
4. If dataset $j$ was stopped early for futility then we do not reject the null hypothesis. If dataset $j$ reached full enrollment, we reject the null hypothesis $H_0: p \leq p_0$ at posterior threshold $k$ if $PPP_{lk} > \theta_k$.

The function returns a list, the first element of which is a `tibble` containing the posterior threshold $\theta$, the predictive threshold $\theta^*$, the mean sample size under the null and the alternative, the proportion of positive trials under the null and alternative, and the proportion of trials stopped early under the null and alternative. The proportion of trials simulated under the null hypothesis for which the null hypothesis was rejected is an estimate of the type I error, and the proportion of trials simulated under the alternative hypothesis for which the null hypothesis was rejected is an estimate of the power. The `print()` option will print the results summary for each combination of thresholds, filtered by an acceptable range of type I error and minimum power, if desired. Note that the results will be sensitive to the choice of `nsim` and `S`. We have set what we believe are reasonable defaults and would caution users against reducing these values without careful consideration.

## Design selection

After obtaining results for all combinations of evaluated posterior and predictive thresholds, the next step is to select the ideal design from among the various options. The \CRANpkg{ppseq} package introduces two criteria to assist users in making a selection. The first, called the "optimal accuracy" design, identifies the design that minimizes the Euclidean distance to 0 type I error probability and a power of 1. To accomplish this, the accuracy Euclidean distance (AED) for the design with posterior threshold $k$ and predictive threshold $m$ is calculated as $AED_{km} = w_\alpha * (\alpha_{km} - 0)^2 + w_(1-\beta) * ((1-\beta)_{km}-1)^2$, where $w_\alpha$ and $w_{(1-\beta)}$ are optional weights on the type I error and power, respectively, and $\alpha_{km}$ denotes the estimated type I error and $(1-\beta)_{km}$ denotes the estimated power. The design with the smallest value of $AED_{km}$ is selected as optimal. The second criteria, called the "optimal efficiency" design, identifies the design that minimizes the Euclidean distance to minimal average sample size under the null and maximal average sample size under the alternative. To accomplish this, the efficiency Euclidean distant (EED) for the design with posterior threshold $k$ and predictive threshold $m$ is calculated as $EED_{km} = w_{\bar{N}_{H_0}} * (\bar{N}_{H_0km} - min(\bar{N}_{H_0}))^2 + w_{\bar{N}_{H_1}} * (\bar{N}_{H_1km}-max(\bar{N}_{H_1}))^2$, where $w_{\bar{N}_{H_0}}$ and $w_{\bar{N}_{H_1}}$ are optional weights on the average sample size under the null and alternative, respectively, $\bar{N}_{H_0km}$ and $\bar{N}_{H_1km}$ denote the average sample sizes under the null and alternative, respectively, and $min(\bar{N}_{H_0})$ and $max(\bar{N}_{H_1})$ denote the minimum average sample size under the null and the maximum average sample size under the alternative alternative, respectively, across all combinations of $k$ and $m$. The design with the smallest value of $EED_{km}$ is selected as optimal. The `optimize_design()` function returns a list that contains the details of each of the two optimal designs.

@@

future::plan(future::multicore(workers = 40))
